Glyph-maps for Visually Exploring Temporal Patterns in Climate Data and Models

### 2012-03-05 revision/paper.tex
finished perception

Climate data is composed of multiple variables, such as temperature, ozone, precipitation, and winds, with a spatial and temporal context. The classic display for data of this type of is ``heatmap'', a tiled plot using color to display the value of the variable of interest at each location. When measurements are made at multiple time points it is reasonably common to display the data with small multiples~\citep{tufte:2001,IPCC}, such as in Figure~\ref{fig:nasa-facet}. In this plot, a separate map is drawn for each month (columns) and year (rows), over a six year period of remotely sensed temperature data above Central America \citep{murrell:2010}. Color is used to display de-seasonalized temperature, with red mapped to high values and blue to low. The most noticeable feature is the strong red patch in the equatorial Pacific beginning mid-1997 and tapering out during 1998. This is the El Ni\~no event, a major temperature anomaly (see for example, \citet{trenberth} or \url{http://www.esrl.noaa.gov/psd/enso/enso.description.html}.) More work is required to see more localized patterns, like the cooler land temperatures seen in early years.

@@

The glyph-map can also be thought of as a small multiple display. The primary comparison is different from the facetted map. Cognitively, elements placed close to each other are more easily and accurately compared \citep{cleveland:1984}. The facetted colored map, places greater emphasis on comparison between the spatial locations, because these are positioned together, and secondary emphasis on time because these are separated by facets. In the glyph-map, primary emphasis is placed on the temporal comparison, and secondarily on spatial comparison, because facetting is on spatial locations.

@@

\begin{figure*}[htbp]
  \centering
  \includegraphics[width=5.5in]{nasa-colored-map.png}
  \caption{Facetted map of de-seasonalized temperature. The dominant feature is the El Ni\~no warming in the southern equatorial region in the last half of 1997 and first half of 1998. Smaller features are only noticeable on closer inspection, or if pointed out: such as the relative warming on the mountain regions in south and north America in the later years.}
  \label{fig:nasa-facet}
\end{figure*}

@@

@@

%Glyph-maps can be generated with existing graphics software after performing a simple pre-processing step, making them easily accessible to a wide audience. 
To create a glyph-map it is useful to recognize that the glyph-map are a linear mix of two structural components of the data: spatial location and data values. The spatial location is the major positioning component, while the data values are minor adjustments to those positions. For spatiotemporal data, the major axes are latitude ($y_{\amaj}$) and longitude ($x_{\amaj}$), and the minor axes are time ($x_{\amin}$) and some measurement ($y_{\amin}$), for example, temperature, or predicted temperature. Assuming the minor axes are rescaled to $[-1, 1]$, the final coordinates $(x,y)$ on the chart are the linear combination, given by:

@@

\noindent where $w$ and $h$ are maximal width and height of the glyph, respectively. For gridded data, it is convenient to make use of the resolution of the major axes, that is, the minimum difference between neighboring locations.

Since the coordinates are linear combinations of two variables, it is possible to build these these line-glyphs interactively, illustrated by software packages that have incorporated tours (as described in \citet{cook:2006}), such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or GGobi \citep{swayne:2003}. This technique is used to good effect in \citet{buja:1996a}, and was used to examine the climate data in \citet{hobbs:2010}.

@@

Figure~\ref{fig:templates} gives some small examples of time series with prototypical trends, shown both in cartesian coordinates and polar coordinates. Differences between linear and nonlinear trend are more apparent in the line-glyphs, and are effectively lost in the star-glyphs, which are most effective in exposing cyclical patterns due to seasonality. Star-glyphs show seasonality as floral cartoons, with peaks forming petals. It is surprising that time series with opposing trends in line-glyphs don't show this symmetry when displayed with star-glyphs. The area of star-glyphs coordinates mainly reflect the average value of a time series, while its shape shows deviations from the average.

@@

Glyph-maps allow time trends to be read directly from the plot. Values are mapped to position, one of the easiest properties to perceive \citep{cleveland:1984}. Perceiving time trends in faceted heatmaps is much more difficult: not only do you need to read value from color (a relatively difficult task), you also need to spot the difference between the different maps. This is challenging from a cognitive perspective \citep{healey:2011,busey} and comparisons can suffer from \emph{change blindness}, often leading to failure to notice the change and not knowing that something changed. To see this, examine Figures~\ref{fig:nasa-facet} and \ref{fig:nasa-glyph}. The heatmap requires comparing color values across maps to assess temporal change but the glyph-map allows direct reading of the temporal trend at each location, and comparison across spatial neighbors.

@@

One draw-back of the glyph-maps, particularly if linear trend is displayed as an icon, is they may suffer from the Z\"ollner Illusion \citep{Zollner}, visible in Figure \ref{fig:gistemp-pred}, which makes straight lines look crooked.

A third alternative is to calculate the relevant statistic for each location and display these values as single heatmap. For example, to study long-term trends, we could calculate the slope of a linear model at each location and color tiles across the map by the value of the slope. The disadvantage of this approach is that it is not exploratory, and requires that we have a very precise definition of the measurement we are interested in, and it still suffers from the relatively difficult color perception.

@@

Climate data is usually large: many locations in space, and many points in time. This makes climate data arduous to work with computationally, and to visualize. Careful use of models to decompose data into long-term trend, seasonal effects and error is vital for to make the problem both computationally and visually tractable.

>>>>

Climate data is composed of multiple variables, such as temperature, ozone, precipitation, and winds, with a spatial and temporal context. The classic display for data of this type is a ``heatmap'', a tiled plot using color to display the value of the variable of interest at each location. When measurements are made at multiple time points it is reasonably common to display the data with small multiples~\citep{tufte:2001,IPCC}, such as in Figure~\ref{fig:nasa-facet}. In this plot, a separate map is drawn for each month (columns) and year (rows), over a six year period of remotely sensed temperature data above Central America \citep{murrell:2010}. Color is used to display de-seasonalized temperature, with red mapped to high values and blue to low. The most noticeable feature is the strong red patch in the equatorial Pacific beginning mid-1997 and tapering out during 1998. This is the El Ni\~no event, a major temperature anomaly (see for example, \citet{trenberth} or \url{http://www.esrl.noaa.gov/psd/enso/enso.description.html}.) More work is required to see more localized patterns, like the cooler land temperatures seen in early years.

@@

The glyph-map can also be thought of as a small multiple display. The primary comparison is different from the facetted heatmap. Cognitively, elements placed close to each other are more easily and accurately compared \citep{cleveland:1984}. The facetted heatmap, places greater emphasis on comparison between the spatial locations, because these are positioned together, and secondary emphasis on time because these are separated by facets. In the glyph-map, primary emphasis is placed on the temporal comparison, and secondarily on spatial comparison, because facetting is on spatial locations.

@@

\begin{figure*}[htbp]
  \centering
  \includegraphics[width=5.5in]{nasa-colored-map.png}
  \caption{Facetted heatmap of de-seasonalized temperature. The dominant feature is the El Ni\~no warming in the southern equatorial region in the last half of 1997 and first half of 1998. Smaller features are only noticeable on closer inspection, or if pointed out: such as the relative warming on the mountain regions in south and north America in the later years.}
  \label{fig:nasa-facet}
\end{figure*}

@@

Another common approach for displaying spatiotemporal data is to calculate a statistical summary for each location and display these values as single heatmap. For example, to study long-term trends, the slope of a linear model can be calculated at each location and color tiles used to display the slope value over space. This approach is less than exploratory, in the sense that it requires a pre-conceived definition of the relationship of interest, and clean data so that the summary statistic adequately summarizes the pattern.

@@

%Glyph-maps can be generated with existing graphics software after performing a simple pre-processing step, making them easily accessible to a wide audience. 
To create a glyph-map it is useful to recognize that the glyph-map is a linear mix of two structural components of the data: spatial location and data values. The spatial location is the major positioning component, while the data values are minor adjustments to those positions. For spatiotemporal data, the major components are latitude ($y_{\amaj}$) and longitude ($x_{\amaj}$), and the minor components are time ($x_{\amin}$) and some measurement ($y_{\amin}$), for example, temperature, or predicted temperature. Assuming the minor components are rescaled to $[-1, 1]$, the final coordinates $(x,y)$ on the chart are the linear combination, given by:

@@

\noindent where $w$ and $h$ are scale parameters, setting the width and height of the glyph, respectively. For gridded data, $w$ and $h$ will be the minimum difference between neighboring spatial locations to make use of the resolution of the major components.

This way of thinking of the glyph-map is convenenient for several reasons. The first is that it moves the plots beyond the small multiples, many little plots, approach. That provides flexibility to cover large spatial domains using potentially tiny glyphs. A second, and major, reason is that with this approach they can then be readily made to be interactive. Different variables can be swept into, and out of, the minor axes using mouse motion by varying $w$ and $h$ proportional to mouse movement. Interactive glyph-maps were used to explore the data-expo data using  GGobi \citep{swayne:2003}, and then later reproduced in publication quality resolution using {\tt ggplot2} \citep{me:ggplot2} in \citet{hobbs:2010}. In GGobi, a manual tour is used to make the linear combination of spatial variables with other variables \citep{CB95}, and equation \ref{coords.eqn} makes this approach more explicit. The third reason is that it enables adjusting scales of glyphs in multiple ways to vary the emphasis, which is discussed in Section \ref{sec:perception}.

%Since the coordinates are linear combinations of two variables, it is possible to build these these line-glyphs interactively, illustrated by software packages that have incorporated tours (as described in \citet{cook:2006}), such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or. This technique is used to good effect in \citet{buja:1996a}, and was used to examine the climate data in \citet{hobbs:2010}.

@@

Figure~\ref{fig:templates} gives some small examples of time series with prototypical trends, shown both in Cartesian coordinates and polar coordinates. Differences between linear and nonlinear trend are more apparent in the line-glyphs, and are effectively lost in the star-glyphs, which are most effective in exposing cyclical patterns due to seasonality. Star-glyphs show seasonality as floral cartoons, with peaks forming petals. It is surprising that time series with opposing trends in line-glyphs don't show this symmetry when displayed with star-glyphs. The area of star-glyphs coordinates mainly reflect the average value of a time series, while its shape shows deviations from the average.

@@

Glyph-maps allow time trends to be read directly from the plot. Values are mapped to position, one of the easiest properties to perceive, rather than color, which is among the hardest graphical element to parse \citep{cleveland:1984}. Perceiving time trends in faceted heatmaps is much more difficult: not only do you need to read value from color (difficult), you also need to spot the difference (major challenge) between the different maps. From a cognitive perspective these sorts of comparisons can suffer from \emph{change blindness}, often leading to failure to notice the change and not knowing that something changed \citep{healey:2011,busey}. To see this, examine Figures~\ref{fig:nasa-facet} and \ref{fig:nasa-glyph}. The heatmap requires comparing color values across maps to assess temporal change but the glyph-map allows direct reading of the temporal trend at each location, and comparison across spatial neighbors. Conversely, if the purpose is to read spatial trend for one time point the heatmap may be a better choice, because this is a difficult task with the glyph-map.

@@

A possible draw-back of the glyph-map, particularly if linear trend is displayed as an icon, is it may suffer from the Z\"ollner Illusion \citep{Zollner}, visible in Figure \ref{fig:gistemp-pred}, which makes straight lines look crooked.

@@

Climate data is usually large: many locations in space, and many points in time. This makes climate data arduous to work with computationally, and to visualize. There is a limit to the resolution of human visual system \citep{Kr12} which will prevent perfect perception of large spatial domains or seasonality of many years of data. Exploring large data requires reducing it to a resolution that can be reasonably absorbed, and statistical models can be useful for this.
Models can be used to decompose data into long-term trend, seasonal effects and error, which is vital for making the problem both computationally and visually tractable.
